---
title: "Data Requirements for <em>autospc</em> Analysis"
output:
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
    fig_caption: yes
    toc: true
    number_sections: true
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Data Requirements for <em>autospc</em> Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=7,
  fig.height=5
)
```

```{r setup, message=FALSE}
library(autospc)
library(dplyr)
```

For each type of Statistical Process Control (SPC) chart supported by *autospc*,
certain data columns must be specified, and those columns must meet certain
requirements. This article sets out these requirements for each chart type.

There are two factors influencing the nature of these requirements. First, the
statistical theory behind SPC charts places restrictions on the type of data
each chart supports. For instance, the theory behind C-charts draws on the
Poisson distribution, which is a discrete distribution defined on the
non-negative integers. Therefore the variable of interest for a C-chart must be
a non-negative integer ( \(0, 1, 2, \dots\)). Second, the types of objects used
to represent data in **R**. For instance, a non-negative integer is most
naturally represented as an object of type "`integer`", but there may also be
situations in which it is appropriate to use an object of type "`double`".

The type requirements of *autospc* are intended to be a flexible as possible,
whilst reflecting the natural usage of the **R**'s object types. For example,
whilst technically an object of type "`integer`" can be coerced to type
"`logical`", *autospc* does not allow "`integer`" data to be used in place of
"`logical`" in observation-level data for a P-chart, since the risk of confusion
with aggregated data is too great.

Below is an example call to `autospc()`. Note that as is often the case, the
first argument (`data`) is passed by position, and the rest are passed as named
arguments. 

```{r example-xmr, fig.cap="Example XMR chart", fig.show='hide', message=FALSE}
autospc(
  ed_attendances_monthly, 
  chart_type = "C'", 
  x = month_start, 
  y = att_all
)
```

In what follows, we shall refer to the columns of the passed data (in this
example, the data `ed_attendances_monthly`) by the names of the arguments they
are passed to. So in this example, in the sentence "The `y` column must be
either an '`integer`' or '`double`'", what is really meant is "The `att_all`
column must be either an '`integer`' or '`double`'".

# The subgrouping variable, `x`

The column specified as `x` will be used both in aggregating data into subgroups
for plotting (where this is relevant), and as the variable plotted on the
horizontal axis (i.e. ordering the data) on the chart. Object types that are
currently supported for `x` are: "`Date`", "`POSIXct`", "`double`", and
"`integer`".

# XMR charts

The data columns required for the XMR chart are as follows: 

* The subgrouping variable, to be plotted on the horizontal axis, `x`
* The variable of interest, to be plotted on the vertical axis, `y`. This must
be of type "`integer`" or "`double`".







