---
title: "Data Requirements for autospc Analysis"
output:
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
    fig_caption: yes
    toc: true
    number_sections: true
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Data Requirements for autospc Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup, message=FALSE}
library(autospc)
```

For each type of Statistical Process Control (SPC) chart supported by *autospc*,
certain data columns must be specified, and those columns must meet certain
requirements. This article sets out these requirements for each chart type.

There are two factors influencing the nature of these requirements. First, the
statistical theory behind SPC charts places restrictions on the type of data
each chart supports. For instance, the theory behind C-charts draws on the
Poisson distribution, which is a discrete distribution defined on the
non-negative integers. Therefore the variable of interest for a C-chart must be
a non-negative integer ( \(0, 1, 2, \dots\)). Second, the types of objects used
to represent data in **R**. For instance, a non-negative integer is most
naturally represented as an object of type `integer`, but there may also be
situations in which it is appropriate to use an object of type `double`.

The type requirements of *autospc* are intended to be a flexible as possible,
whilst reflecting the natural usage of the **R**'s object types. For example,
whilst technically an object of type `integer` can be coerced to type
`logical`, *autospc* does not allow `integer` data to be used in place of
`logical` in observation-level data for a P-chart, since the risk of confusion
with aggregated data is too great.

Below is an example call to `autospc()`. Note that as is often the case, the
first argument (`data`) is passed by position, and the rest are passed as named
arguments. 

```{r example-c-prime, fig.cap="Example C' chart", fig.show='hide', message=FALSE}
autospc(
  ed_attendances_monthly,
  chart_type = "C'",
  x = month_start,
  y = att_all
)
```

In what follows, we shall refer to the columns of the passed data (in this
example, the data `ed_attendances_monthly`) by the names of the arguments they
are passed to. So in this example, in the sentence "The `y` column must be
either an `integer` or `double`", what is really meant is "The `att_all`
column must be either an `integer` or `double`".

# The subgrouping variable, `x`

The column specified as `x` will be used both in aggregating data into subgroups
for plotting (where this is relevant), and as the variable plotted on the
horizontal axis (i.e. ordering the data) on the chart. Object classes that are
currently supported for `x` are: `Date`, `POSIXct`, `numeric`, and
`integer`.

# XMR charts (and their components, X and MR charts)

The data columns required for XMR, X and MR charts are as follows: 

* The subgrouping variable, to be plotted on the horizontal axis, `x`
* The variable of interest, to be plotted on the vertical axis, `y`. This must
be of type `integer` or `double`.

Unlike C/C' charts (see Section \@ref(c-and-c-charts)), XMR charts place no
further restriction on `y`: non-whole-number doubles are accepted without
modification or warning, since they are suitable for continuous measurements
as well as counts.

# XbarS charts (and their components, Xbar and S charts)

Xbar and S charts are for measurements taken in subgroups. The Xbar chart plots
each subgroup's mean and the S chart each subgroup's sample standard deviation.
For example,
[Adeleke et al. (2019)](https://doi.org/10.1136/openhrt-2019-001086) used an
XbarS chart as a balancing measure in an initiative to improve the care of
patients with atrial fibrillation (AF) in 48 general practices in Hounslow,
London. Each subgroup was a month, and each measurement the CHA~2~DS~2~-VASc
stroke risk score of a patient newly diagnosed with AF in that month, so the
chart showed whether the stroke risk of the patients being diagnosed changed as
the initiative went on. Its supplementary file 6 shows the Xbar chart, whose
limits vary from month to month with the number of patients diagnosed.

*autospc* supports two ways of supplying the data: observation-level data, with
one row per measurement, such as one row per newly diagnosed patient; or
aggregated data, with each row giving the mean, number and standard deviation of
some measurements, such as one row per practice per month. In both cases the
`x` column is required as for the above chart types, and rows that share a
value of `x` form one subgroup, so `x` may be repeated.

## Observation-level data (no `n` or `s` specified)

If neither `n` nor `s` is specified, each row is one measurement, and `y` is
the measurement. `y` must be of type `integer` or `double`, and *autospc*
computes each subgroup's mean, size and standard deviation from the rows that
share its `x`.

## Aggregated data (`n` and `s` specified)

If `n` and `s` are specified, each row summarises some measurements: `y` is
their mean, `n` their number and `s` their sample standard deviation. All three
must be of type `integer` or `double`. Where several rows share an `x`, as
where each row holds one practice's patients in a month, they are combined into
one subgroup whose mean, size and standard deviation are those of all their
measurements together.

The two must be specified together. Specifying only one causes an error:

> For Xbar, S and XbarS charts given one row per subgroup, n and s must both
> be specified: y is then the subgroup mean, n its size and s its sample
> standard deviation.

A column called `n` or `s` in the data is used for that argument without being
specified, as `n` is for P and P' charts.

Neither `n` nor `s` may be negative:

> For Xbar, S and XbarS charts, n and s cannot be negative.

The subgroup size `n` must be a whole number. Where it is of type `double`
with at least one non-whole-number value, the values are rounded to the
nearest whole number and a warning is issued:

> At least one element of n has non-zero fractional part. Rounding to the
> nearest whole number.
> Xbar, S and XbarS charts require n to be a subgroup size, i.e. whole
> numbers only.

## Subgroups of one

A subgroup of one has a mean but no standard deviation. It is plotted on the
Xbar chart and counts towards its centre line, but has no control limits there.
It is not plotted on the S chart, and has no control limits there either.

An aggregated row with a mean and a size of two or more, but no standard
deviation, is plotted on the Xbar chart with control limits, and is not plotted
on the S chart.

# C and C' charts

The data columns required for C and C' charts are as follows:

* The subgrouping variable, to be plotted on the horizontal axis, `x`
* The count of events, to be plotted on the vertical axis, `y`. This must be of
type `integer` or `double`.

Since C and C' charts are for count data, `y` must consist of whole numbers
only. *autospc* handles two R types that can represent whole numbers
differently:

* **`integer`**: accepted without modification or warning.
* **`double`**: *autospc* checks whether all values of `y` are whole numbers
(up to machine precision). If they are, the column is accepted without
modification or warning. If at least one value has a non-zero fractional
part, the values are rounded to the nearest whole number and a warning is
issued:

  > At least one element of y has non-zero fractional part. Rounding to
  > the nearest whole number. C and C' charts are for count data, i.e.
  > whole numbers only.

Any other type for `y`, including `logical`, will cause an error:

> For a C or C' chart, y must be of type integer or double.

# P and P' charts

P and P' charts are for proportions. They require a numerator (the count meeting
some criterion) and a denominator (the total count). *autospc* supports two ways
of supplying this information, depending on the form in which the data are
available: observation-level data using a `logical` `y` column, or aggregated
data using numeric `y` and `n` columns. In both cases the `x` column is required
as for the above chart types.

## Observation-level data (no `n` specified)

If `n` is not specified, *autospc* expects `y` to be a column of type `logical`,
where each row represents an individual observation and the value of `y`
indicates whether that observation meets the criterion of interest (e.g. `TRUE`
if a patient attending an emergency department was discharged, admitted or
transferred within 4 hours, `FALSE` otherwise). In this case, *autospc*
internally computes the numerator and denominator by aggregating over subgroups
defined by `x`.

Any type for `y` other than `logical` when `n` is absent will cause an error:

> n is not specified and y is not of type logical. For P and P' charts, if n
> is not specified, y must be of type logical.

## Aggregated data (`n` specified)

If `n` is specified, *autospc* expects both `y` (the numerator) and `n` (the
denominator) to be counts, i.e. whole numbers. Both columns must be of type
`integer` or `double`, and the same whole-number checking logic described
for C/C' charts in Section \@ref(c-and-c-charts) applies independently to each:

* **`integer`**: accepted without modification or warning.
* **`double` with all whole-number values**: accepted without modification or
warning.
* **`double` with at least one non-whole-number value**: values are rounded to
the nearest whole number and a warning is issued. For `y` the warning reads:

  > At least one element of y has non-zero fractional part. Rounding to the
  > nearest whole number. P and P' charts with n specified require y to be a
  > count, i.e. whole numbers only.

  For `n` the analogous warning reads:

  > At least one element of n has non-zero fractional part. Rounding to the
  > nearest whole number. P and P' charts with n specified require n to be a
  > count, i.e. whole numbers only.

Any other type for `y` when `n` is present, including `logical`, will cause
an error:

> For a P or P' chart with n specified, y must be of type integer or double.

Similarly, any type for `n` other than `integer` or `double` will cause an
error:

> For a P or P' chart with n specified, n must be of type integer or double.

# Summary

Table \@ref(tab:summary-table) summarises the column requirements for each chart
type.

```{r summary-table, echo=FALSE}
summary_df <- data.frame(
  `Chart type` = c(
    "X / MR / XMR",
    "Xbar / S / XbarS (observation level)",
    "Xbar / S / XbarS (aggregated)",
    "C / C'",
    "P / P' (observation level)",
    "P / P' (aggregated)"
  ),
  `y type(s) accepted` = c(
    "`integer`, `double`",
    "`integer`, `double`",
    "`integer`, `double` (the subgroup mean)",
    paste(
      "`integer`, `double` (whole numbers only; non-integer doubles are",
      "rounded with a warning)"
    ),
    "`logical`",
    paste(
      "`integer`, `double` (whole numbers only; non-integer doubles are",
      "rounded with a warning)"
    )
  ),
  `n type(s) accepted` = c(
    "not used",
    "not used",
    paste(
      "`integer`, `double` (the subgroup size; whole numbers only;",
      "non-integer doubles are rounded with a warning)"
    ),
    "not used",
    "not used",
    paste(
      "`integer`, `double` (whole numbers only; non-integer doubles are",
      "rounded with a warning)"
    )
  ),
  `s type(s) accepted` = c(
    "not used",
    "not used",
    "`integer`, `double` (the subgroup's sample standard deviation)",
    "not used",
    "not used",
    "not used"
  ),
  check.names = FALSE
)

knitr::kable(
  summary_df,
  caption = "Summary of data column requirements by chart type."
)
```
